$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-04-05 qdpm_paper.tex
Worked on manuscript

\begin{abstract}
We introduce the quantum dimer-pentamer model (QDPM) on the square lattice. This model is a generalization of the square lattice quantum dimer model (QDM) as its configuration space comprises fully-packed hard-core dimer coverings as well as dimer configurations containing pentamers, where four dimers touch a vertex. Thus in the QDPM, the fully-packed, hard-core constraint of the QDM is relaxed such that the local dimer number at each vertex is fixed modulo 3; correspondingly, the local $U(1)$ gauge symmetry of the QDM Hilbert space is reduced to a local $Z_3$ gauge symmetry in the QDPM. We construct a local Hamiltonian for which the Rokhsar-Kivelson (RK) equal superposition state is the exact ground state and has a 9-fold topological degeneracy on the torus. Using Monte Carlo calculations, we find no spontaneous symmetry breaking in the RK wavefunction and that its dimer-dimer correlation function decays exponentially. By doping the QDPM RK state with a pair of monomers, we demonstrate that $Z_3$ electric charges are deconfined. Additionally, we introduce a $Z_3$ magnetic string operator which shows no signatures of magnetic vortex condensation and with correlations that decay exponentially. These numerical results suggest that the ground state of the QDPM is a dimer liquid with $Z_3$ topological order.
\end{abstract}

@@

Strongly interacting quantum many-body systems are known to support a variety of exotic quantum phases of matter with no classical analogue. One class of such quantum phases of matter are topologically ordered quantum liquids; their non-trivial topological order can not be identified by a local order parameter as these phases of matter don't break conventional symmetries~\cite{Wen1990,Nayak2008}. Both because of of the exotic nature of topological phases and their potential use for quantum information processing~\cite{Freedman2001,Kitaev2003}, there is great interest in the development of theoretical models which display these phases and potentially could provide candidate experimental systems to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}.

A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}; such models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally. Among the simplest class of models displaying topological quantum liquid phases are those with local geometric constraints; the constraints inhibit the formation of symmetry breaking local order and given sufficient dynamics, may lead to a symmetric liquid ground state. One of the most basic of these is the quantum dimer model (QDM) describing dimer degrees of freedom living on the links of a lattice~\cite{Rokhsar1988,Moessner2008}; the hard-core, fully-packed QDM has a local constraint that requires exactly one dimer touching each vertex.

@@

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can be enforced by a local potential energy penalty and the minimal dynamics often arise at the lowest non-trivial order in perturbation theory from a related spin model. We will refer to such models with dimer degrees of freedom on links and local constraints as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right (left) with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $\tilde{n}_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_0$ ($-c_0$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

@@

On surfaces with non-trivial topology, we may define the topological loop operators
\begin{equation}
W_{X,Y} \left( \alpha \right) \equiv S_{\Gamma_{X,Y}} \left( \alpha \right)
\end{equation}
where $\Gamma_{X,Y}$ are topologically non-trivial closed loops that wind once about the $X$ and $Y$ directions of the torus, respectively. Since $W_{X,Y} (\alpha)$ commute with the Hamiltonian, we may use their eigenvalues to label the ground states.

@@

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes $\bplqt$ of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0.
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu $:
\begin{equation}
 {h}_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr)\biggl( \ebra{ c_\nu} - \ebra{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice~\cite{????}, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

Defects in these models violate the local dimer number constraint and thus carry non-zero U(1) electric charge. \tabref{tab:defects} lists all the possible defects and their associated U(1) charges in QDM and QFPLM. When a defect lives on sublattice A, it has the opposite charge that the same defect on sublattice B has. Note that in the QDM, the maximum charge is $\pm 3$, and there are six distinct possible non-trivial electric charges of defects.

The QDM and QFPLM on the square lattice have isolated liquid ground states at the RK point surrounded by symmetry broken phases~\cite{Rokhsar1988,Leung1996,Syljuasen2005,Syljuasen2006,Syljuasen2004,Moessner2008}. The RK point of these models have been shown to have power-law decaying dimer-dimer and monomer-monomer correlation functions and gapless excitations~\cite{Fisher1961,Kasteleyn1961,Fisher1966a,Fisher1963,Kasteleyn1963a,Sutherland1988b,Sutherland1988a,Kohmoto1988a,Liang1988,Sutherland1988,Kohmoto1988,Levitov1990,Leung1996,Henley1997,Henley1997,Henley2004a,Pollmann2006}. The algebraic decay of monomer correlations is indicative of the logarithmic confinement of electric charges in these models. Thus, such fixed number constrained models on bipartite lattices are expected to have an isolated unstable gapless critical quantum liquid points with logarithmically confined defect and an extensive topological degeneracy. These features of fixed number constrained models are consistent with the effective U(1) gauge theory description of the RK point~\cite{Fradkin2004,Fradkin1990,Moessner2001}; in 2+1D pure U(1) gauge theories are confining and thus the lack of an extended deconfined topologically ordered phase near the RK point~\cite{PolyakovBook}.

@@

The additional dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{Z_2} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{Z_2} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{Z_2}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega= (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_3way.pdf}
        \caption
        {
            (a) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}}
            n_{\bm{r}}\rangle$ on an $L=32$ lattice. The origin corresponds to the orange link, and
            $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off
            the given scale. 
            (b) The horizontal dimer density structure factor $S_d(\bm{k})$ for the
            QDPM RK ground state on an $L=64$ lattice. The lack of $x\leftrightarrow y$ rotational
            symmetry is due to the fact that horizontal links have been isolated. 
            (c) The horizontal dimer-dimer structure factor $S_d(\bm{k})$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:dimer_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\columnwidth]{peak_scaling.pdf}
    \caption{Finite-size scaling of the maxima in the dimer ($S_d$) and magnetic string ($S_M$) structure factors. All maxima vanish in the thermodynamic limit, indicating that the RK state of the QDPM is a symmetric dimer liquid. }
    \label{fig:bragg_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Future work could directly probe the existence of a finite gap above the RK point by studying the imaginary time dynamics of correlations functions using Monte Carlo methods similar to those used in here~\cite{Henley1997,Henley2004a}; such an approach could be used to explore if the low energy excitations are best described by $Z_3$ magnetic vorticies~\cite{Ivanov2004,Ralko2007,Misguich2008d}. Additionally, one could conclusively demonstrate the existence of $Z_3$ topological order by computing the topological entanglement entropy in the QDPM ground state Monte Carlo methods~\cite{Levin2005a,Kitaev2006b,Hastings2010}.

@@

We can define two oriented open string operators which commute with $H_{Z_3}$ everywhere except at the endpoints. The electric string operator $S_e$, which acts on an open, oriented string on the lattice $\gamma_{vv'}$ with endpoints $v,v'$, and creates a $+1$ ($-1$) charge at vertex $v$ ($v')$, is defined as
\begin{equation}
    S_e \left( \gamma_{vv'} \right) \equiv \prod_{l \in \gamma_{vv'}} Q_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ in $\gamma_{vv'}$, $v_l$ and $v_l'$ are the two vertices in $l$, ordered according to the orientation of $\gamma_{vv'}$ (see \figref{fig:example_elec_string}). The magnetic string operator $S_m$ is defined by an open, oriented string on the dual lattice and creates a magnetic vortex-antivortex pair on plaquettes at the endpoints of $\tilde{\gamma}_{pp'}$
\begin{equation}
    S_m \left( \tilde{\gamma}_{vv'} \right) \equiv \prod_{l \perp \tilde{\gamma}_{pp'}} P_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ that cross $\tilde{\gamma}$ and $v_l$ ($v_l'$) is the vertex to the left (right) of the string as $\tilde{\gamma}$ is traversed from $p$ to $p'$. 
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\linewidth]{example_elec_string.pdf}
    \caption{ ({\it left}) An example of a charge free $Z_3$ electric string configuration. The electric string operator $S_\gamma^e$ is represented by the thick gray line, oriented from the white to black vertex. ({\it right}) An example of a $Z_3$ electric string configuration with a $+1$ ($-1$) electric charge at the red (blue) vertex; this string configuration results form the action of $S_\gamma^e$ on the left panel.  }
    \label{fig:example_elec_string}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}
%    \centering
%    \includegraphics[width=1.0\columnwidth]{spatial_cors_perp.pdf}
%    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
%    sub-lattice for a $L=64$ lattice.}
%    \label{fig:spatial_dmr_cor}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Additionally, these string operators give rise to a topological degeneracy in systems of non-trivial topology. For example, consider the charge free sector on a torus. We can consider two topologically non-trivial closed loops $\tilde{\Gamma}_{x}$ \& $\tilde{\Gamma}_{Y}$ which have a single winding around the $x$ or $y$ directions of the torus. Since $S_m(\tilde{\Gamma}_{X})$ commutes with $H_{Z_3}$, we can label the energy eigenstates by the eigenvalues of $S_m(\tilde{\Gamma}_{X})$, and similarly for $\tilde{\Gamma}_Y$: $\{1,e^{ i 2\pi/3},e^{- i 2\pi/3}\}$. Thus $H_{Z_3}$ has a 9-fold ground state degeneracy on the torus, and the ground states subspace can be labeled by the $Z_3$ electric flux winding about each axis of the torus.

>>>>

\begin{abstract}
We introduce the quantum dimer-pentamer model (QDPM) on the square lattice. This model is a generalization of the square lattice quantum dimer model as its configuration space comprises fully-packed hard-core dimer coverings as well as dimer configurations containing pentamers, where four dimers touch a vertex. Thus in the QDPM, the fully-packed, hard-core constraint of the quantum dimer model is relaxed such that the local dimer number at each vertex is fixed modulo 3, resulting in an exact local $Z_3$ gauge symmetry. We construct a local Hamiltonian for which the Rokhsar-Kivelson (RK) equal superposition state is the exact ground state and has a 9-fold topological degeneracy on the torus. Using Monte Carlo calculations, we find no spontaneous symmetry breaking in the RK wavefunction and that its dimer-dimer correlation function decays exponentially. By doping the QDPM RK state with a pair of monomers, we demonstrate that $Z_3$ electric charges are deconfined. Additionally, we introduce a $Z_3$ magnetic string operator that we find decays exponentially and shows no signatures of magnetic vortex condensation and with correlations. These numerical results suggest that the ground state of the QDPM is a dimer liquid with $Z_3$ topological order.
\end{abstract}

@@

Strongly interacting quantum many-body systems are known to support a variety of exotic quantum phases of matter with no classical analogue. Among these are topologically ordered quantum liquids; their non-trivial topological order can not be identified by a local order parameter as these phases of matter don't break conventional symmetries~\cite{Wen1990,Nayak2008}. Both because of of the exotic nature of topological phases and their potential use for quantum information processing~\cite{Freedman2001,Kitaev2003}, there is great interest in the development of theoretical models which display these phases and could provide candidate experimental systems to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}.

A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}; such models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally. Among the simplest classes of models displaying topological quantum liquid phases are those with local geometric constraints; the constraints inhibit the formation of symmetry breaking local order and given sufficient dynamics, may lead to a symmetric liquid ground state. The quantum dimer model (QDM) describing dimer degrees of freedom living on the links of a lattice is one of the most basic locally constrained models~\cite{Rokhsar1988,Moessner2008}; the hard-core, fully-packed QDM has a local constraint that requires that exactly one dimer touches each vertex.

@@

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can be enforced by a local potential energy penalty and the minimal dynamics often arise at the lowest non-trivial order in perturbation theory from a related spin model\cite{Albuquerque2008}. We will refer to such models with dimer degrees of freedom on links and local constraints as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2015,Buerschaper2014}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right (left) with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $\tilde{n}_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_P$ ($-c_P$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

@@

On surfaces with non-trivial topology, we may define the topological loop operators
\begin{equation}
W_{X} \left( \alpha \right) \equiv S_{\Gamma_{X}} \left( \alpha \right)
\end{equation}
where $\Gamma_{X}$ is a topologically non-trivial closed loop that winds once about the $x$ axis; we define $W_Y$ similarly. Since $W_{X} (\alpha)$ and $W_{Y} (\alpha)$ commute with the Hamiltonian, we may use their eigenvalues to label the ground states.

@@

The Hamiltonian for these models gives dynamics to parallel dimers on the plaquettes of the lattice:
\begin{equation}
\HQDM\left(t,v\right) \equiv \HGQDM \biggl( \bigl\{ \left(\vrtplqt,\hrzplqt\right) \bigr\},\bigl\{ \left(t,v\right) \bigr\} \biggr)  \label{eq:HQDM}
\end{equation}
where we have chosen $t_\bplqt = t$, $v_\bplqt =v, \forall \bplqt$.
%\begin{align}
%    \label{}
%   \HQDM = &-t\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\hrzplqt} +h.c. \biggr) \notag\\
%    					&+v\sum_{\bplqt} \biggl( \eket{\vrtplqt} \ebra{\vrtplqt} +\eket{\hrzplqt} \ebra{\hrzplqt} \biggr). \label{eq:HQDM}
%\end{align}
In this case of a fixed number constraint, the Hamiltonian conserves $Q_v$ and therefore the dimer flux through any closed loop such that:
\begin{equation}
\left[ \Phi_\Gamma, \HQDM \right] = 0.
\end{equation}
We can define two conserved topological winding numbers $\Phi_{X}$ \& $\Phi_{Y}$ which are the dimer flux through topologically non-trivial loops that wind around each axis of the lattice when the system has periodic boundary conditions (see \figref{fig:orientedSL}). On a lattice of dimensions $L_x\times L_y$, the eigenvalues of $\Phi_X$ and $\Phi_Y$ are integers in the range
\begin{equation}
-L_{x}/2 \leq  \Phi_{X} \leq L_{x}/2,
\end{equation}
and similarly for $\Phi_Y$, thus there are $\mathcal{O} (L_x L_y)$ distinct topological sectors.

At the Rokhsar-Kivelson point $t=v$, we may write $\HQDM$ in the form of \Eqref{eq:HRK} by defining projection operators  $ {h}_\nu $:
\begin{equation}
 {h}_\nu \equiv \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr)\biggl( \ebra{ c_\nu} - \ebra{\bar{c}_\nu} \biggr).
\end{equation} 
Since the plaquette flip term in \Eqref{eq:HQDM} is believed to be ergodic in each of topological sector $(\Phi_x,\Phi_y)$ on the square lattice, each topological sector corresponds to a unique RK ground state where $\Omega$ is labeled by $(\Phi_x,\Phi_y)$. Thus the  RK points of the QDM and QFPLM have an \emph{extensively} degenerate ground states.

Defects in these models violate the local dimer number constraint and thus carry non-zero U(1) electric charge. \tabref{tab:defects} lists all the possible defects and their associated U(1) charges in QDM and QFPLM. When a defect lives on sublattice A, it has the opposite charge that the same defect on sublattice B has. Note that in the QDM, the maximum charge is $\pm 3$.

The QDM and QFPLM on the square lattice have isolated liquid ground states at the RK point surrounded by symmetry broken phases~\cite{Rokhsar1988,Leung1996,Syljuasen2005,Syljuasen2006,Syljuasen2004,Moessner2008}. The RK point of these models have been shown to have power-law decaying dimer-dimer and monomer-monomer correlation functions and gapless excitations~\cite{Fisher1961,Kasteleyn1961,Fisher1966a,Fisher1963,Kasteleyn1963a,Sutherland1988b,Sutherland1988a,Kohmoto1988a,Liang1988,Sutherland1988,Kohmoto1988,Levitov1990,Leung1996,Henley1997,Henley1997,Henley2004a,Pollmann2006}. The algebraic decay of monomer correlations is indicative of the logarithmic confinement of electric charges in these models. Thus, such fixed number constrained models on bipartite lattices are expected to have an isolated unstable gapless critical quantum liquid points with logarithmically confined defect and an extensive topological degeneracy. These features of fixed number constrained models are consistent with the effective U(1) gauge theory description of the RK point~\cite{Fradkin2004,Fradkin1990,Moessner2001}; in 2+1D pure U(1) gauge theories are confining and thus the lack of an extended deconfined topologically ordered phase near the RK point~\cite{Polyakov1987}.

@@

The additional dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{Z_2} \equiv S_\gamma \left( \pi \right), \quad W_{X}^{Z_2} \equiv W_{X} \left( \pi \right), \notag
\end{equation}
and $W_Y^{Z_2}$ is defined similarly. Thus the $W_{X}^{Z_2}$ and $W_{Y}^{Z_2}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega= (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_3way.pdf}
        \caption
        {
            (a) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}}
            n_{\bm{r}}\rangle$ on an $L=32$ lattice. The origin corresponds to the orange link, and
            $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off
            the given scale. 
            (b) The horizontal dimer density structure factor $S_d(\bm{k})$ for the
            QDPM RK ground state on an $L=64$ lattice. The lack of $x\leftrightarrow y$ rotational
            symmetry is due to the fact that horizontal links have been isolated. 
            (c) The horizontal dimer-dimer structure factor $S_d(\bm{k})$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:dimer_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\columnwidth]{peak_scaling.pdf}
    \caption{Finite-size scaling of the maxima in the dimer ($S_d$) and magnetic string ($S_M$) structure factors. The lines are fits to $L^{-2}$ scaling. All maxima vanish in the thermodynamic limit, indicating that the RK state of the QDPM is a symmetric dimer liquid. }
    \label{fig:bragg_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Future work could directly probe the existence of a finite gap above the RK point by studying the imaginary time dynamics of correlations functions using Monte Carlo methods similar to those used in here~\cite{Henley1997,Henley2004a}; such an approach could be used to explore if the low energy excitations are best described by $Z_3$ magnetic vorticies~\cite{Ivanov2004,Ralko2007,Misguich2008d}. Additionally, one could conclusively demonstrate the existence of $Z_3$ topological order by computing the topological entanglement entropy in the QDPM ground state using Monte Carlo methods~\cite{Levin2005a,Kitaev2006b,Hastings2010}.

@@

We can define two oriented open string operators which commute with $H_{Z_3}$ everywhere except at the endpoints. The electric string operator $S_e$, which acts on an open, oriented string on the lattice $\gamma_{vv'}$ with endpoints $v,v'$, and creates a $+1$ ($-1$) charge at vertex $v$ ($v')$, is defined as
\begin{equation}
    S_e \left( \gamma_{vv'} \right) \equiv \prod_{l \in \gamma_{vv'}} Q_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ in $\gamma_{vv'}$, $v_l$ and $v_l'$ are the two vertices in $l$, ordered according to the orientation of $\gamma_{vv'}$ (see \figref{fig:example_elec_string}). The magnetic string operator $S_m$ is defined by an open, oriented string on the dual lattice and creates a magnetic vortex-antivortex pair on plaquettes at the endpoints of $\tilde{\gamma}_{pp'}$
\begin{equation}
    S_m \left( \tilde{\gamma}_{pp'} \right) \equiv \prod_{l \perp \tilde{\gamma}_{pp'}} P_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ that cross $\tilde{\gamma}$ and $v_l$ ($v_l'$) is the vertex to the left (right) of the string as $\tilde{\gamma}$ is traversed from $p$ to $p'$. 
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\linewidth]{example_elec_string.pdf}
    \caption{ ({\it left}) An example of a charge free $Z_3$ electric string configuration. The electric string operator $S_\gamma^e$ is represented by the thick gray line, oriented from the white to black vertex. ({\it right}) An example of a $Z_3$ electric string configuration with a $+1$ ($-1$) electric charge at the red (blue) vertex; this string configuration results form the action of $S_\gamma^e$ on the left panel. The magnetic loop operator $S_m (\tilde{\Gamma})$ can detect the presence of the electric charges.  }
    \label{fig:example_elec_string}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}
%    \centering
%    \includegraphics[width=1.0\columnwidth]{spatial_cors_perp.pdf}
%    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
%    sub-lattice for a $L=64$ lattice.}
%    \label{fig:spatial_dmr_cor}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Additionally, these string operators give rise to a topological degeneracy in systems of non-trivial topology. For example, consider the charge free sector on a torus. We can consider two topologically non-trivial closed loops $\tilde{\Gamma}_{X}$ \& $\tilde{\Gamma}_{Y}$ which have a single winding around the $x$ or $y$ directions of the torus. Since $S_m(\tilde{\Gamma}_{X})$ commutes with $H_{Z_3}$, we can label the energy eigenstates by the eigenvalues of $S_m(\tilde{\Gamma}_{X})$, and similarly for $\tilde{\Gamma}_Y$: $\{1,e^{ i 2\pi/3},e^{- i 2\pi/3}\}$. Thus $H_{Z_3}$ has a 9-fold ground state degeneracy on the torus, and the ground states subspace can be labeled by the $Z_3$ electric flux winding about each axis of the torus.
